**Effective application of machine learning to microbiome-based classification problems**

### 2019-10-21 submission/manuscript.Rmd
last edits before submission

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances. Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a modeling approach has often been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using non-linear models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although in some cases, complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, other approaches may be more appropriate. For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may (but not always) emphasize predictive performance. Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modeling approach.

@@

__Model selection and pipeline construction__. We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].

@@

Model selection requires tuning hyperparameters. Hyperparameters are parameters that need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls. An AUROC value of 0.5 indicates that the model's predictions are no different than random. To select hyperparameters, we performed a grid search for hyperparameter settings when training the models. Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and should to be optimized for each new ML task. For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning [Figure S1].

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2]. The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data. The largest difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and give us confidence in our estimate of the generalization performance of the models.

@@

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. We did not take advantage of parallelization in training, nor was the code optimized for speed. Still, as expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].

@@

We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We used the AUROC metric to evaluate predictive performance as our dataset was highly balanced. We repeated this data-split 100 times to measure the possible variation in predictive performance. During training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline generalized to the held-out test sets.

@@

**16S rRNA gene sequencing data.** Stool samples provided by the patients were used for 16S rRNA gene sequencing to measure bacterial population abundances. The sequence data used in our analyses were originally generated by Baxter et al. (available through NCBI Sequence Read Archive [SRP062005], 2015). The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}). These OTU abundances were the features we used to predict colorectal health of the patients. There were 6920 OTUs. OTU abundances were subsampled to the size of the smallest sample and normalized across samples such that the highest abundance of each OTU would be 1, and the lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. A grid search was needed to avoid large variation in prediction performance. For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization. For SVM with radial basis function kernel, we also tuned the sigma hyperparameter, which determines the reach of a single training instance where, for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the depth of the tree where the deeper the tree, the more splits it has. For random forest, we tuned the number of features to consider when looking for the best tree split. For XGBoost, we tuned the learning rate and the fraction of samples used for fitting the individual base learners. Performing a grid search for hyperparameter selection might not be feasible for when there are more than two hyperparameters to tune for. In such cases, it is more efficient to use random search or recently developed tools such as Hyperband to identify good hyperparameter configurations [@li_hyperband:_2016].

The computational burden during model training due to model complexity was reduced by parallelizing segments of the ML pipeline. We parallelized the training of each data-split. This allowed the 100 data-splits to be processed through the ML pipeline simultaneously at the same time for each model. It is possible to further parallelize the cross-validation step for each hyperparameter setting.

@@

**Acknowledgements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network. We would like to thank the members of the Schloss lab for their valuable feedback. Salary support for M.R came from NIH grant 1R01CA215574. Salary support for P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each data-split. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red. Some of the same OTUs were identified as important in all of the linear models. 
\newpage
\includegraphics{Figure_4.png}

**Figure 5. Training times of seven ML models.** The median training time was the highest for XGBoost and shortest for L2-regularized logistic regression. 
\newpage
\includegraphics{Figure_S1.png}

**Figure S1. Hyperparameter setting performances for linear models.** (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model. The stars represent the highest performing hyperparameter setting for each model. 
\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

\newpage
\captionsetup{labelformat=empty}
\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
\small
\begin{tabular}{|l|l|l|}
\hline

@@

\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices.}
\small
\begin{tabular}{|l|l|l|l|}
\hline

@@

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

@@

>>>>

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances. Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a modeling approach has often been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using non-linear models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although in some cases, complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, other approaches may be more appropriate. For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance. Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modeling approach.

@@

__Model selection and pipeline construction__. We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1, Table 1].

@@

Model selection requires tuning hyperparameters. Hyperparameters are parameters that need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls. An AUROC value of 0.5 indicates that the model's predictions are no different than random. To select hyperparameters, we performed a grid search for hyperparameter settings when training the models. Default hyperparameter settings in developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task. For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning as the default C parameter was 1 [Figure S1].

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2]. The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well as non-linear models.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data. The largest difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences were relatively small and gave us confidence in our estimate of the generalization performance of the models.

@@

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. The training times increased with the complexity of the model and the number of potential hyperparameter combinations. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].

@@

We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We used the AUROC metric to evaluate predictive performance as it is a clinically relevant evaluation metric for our study. We repeated this data-split 100 times to measure the possible variation in predictive performance. During training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline generalized to the held-out test sets.

@@

**16S rRNA gene sequencing data.** Stool samples provided by the patients were used for 16S rRNA gene sequencing to measure bacterial population abundances. The sequence data used in our analyses were originally generated by Baxter et al. (available through NCBI Sequence Read Archive [SRP062005], @baxter_microbiota-based_2016). The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}). These OTU abundances were the features we used to predict colorectal health of the patients. There were 6920 OTUs. OTU abundances were subsampled to the size of the smallest sample and normalized across samples such that the highest abundance of each OTU would be 1, and the lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter.For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization. For SVM with radial basis function kernel, we also tuned the sigma hyperparameter, which determines the reach of a single training instance where, for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the depth of the tree where the deeper the tree, the more splits it has. For random forest, we tuned the number of features to consider when looking for the best tree split. For XGBoost, we tuned the learning rate and the fraction of samples used for fitting the individual base learners. Performing a grid search for hyperparameter selection might not be feasible for when there are more than two hyperparameters to tune for. In such cases, it is more efficient to use random search or recently developed tools such as Hyperband to identify good hyperparameter configurations [@li_hyperband:_2016].

The computational burden during model training due to model complexity was reduced by parallelizing segments of the ML pipeline. We parallelized the training of each data-split. This allowed the 100 data-splits to be processed through the ML pipeline simultaneously at the same time for each model. It is possible to further parallelize the cross-validation step for each hyperparameter setting which we have not performed in this study.

@@

**Acknowledgements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network. We would like to thank the members of the Schloss lab for their valuable feedback. Salary support for MR came from NIH grant 1R01CA215574. Salary support for PDS came from NIH grants P30DK034933 and 1R01CA215574.

\captionsetup{labelformat=empty}
\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
\small
\begin{tabular}{|l|l|l|}
\hline

@@

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each data-split. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red. 
\newpage
\includegraphics{Figure_4.png}

**Figure 5. Training times of seven ML models.** The median training time was the highest for XGBoost and shortest for L2-regularized logistic regression. 
\newpage

\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
\small
\begin{tabular}{|l|l|l|l|}
\hline

@@

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

@@

\includegraphics{Figure_S1.png}
**Figure S1. Hyperparameter setting performances for linear models.** (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model. The stars represent the highest performing hyperparameter setting for each model. 
\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
